Thesis Title

### 2022-03-02 chapters/introduction.tex
remove table - allows compilation

\begin{table*}[h!]
\begin{tabular}{@{}lll@{}}

\caption{\textbf{Phred quality scores are logarithmically linked to error probabilities.}}\label{tab:phred_quality}

\toprule
Phred Quality Score & Probability of incorrect base call & Base call accuracy \\ 
\midrule
10                  & 1 in 10                            & 90\%               \\
20                  & 1 in 100                           & 99\%               \\
30                  & 1 in 1000                          & 99.9\%             \\
40                  & 1 in 10,000                        & 99.99\%            \\
50                  & 1 in 100,000                       & 99.999\%           \\
60                  & 1 in 1,000,000                     & 99.9999\%          \\ \bottomrule
\end{tabular}
\end{table*}

Since their introduction, Phred scores have become the \textit{de facto} standard for representing sequencing read base qualities \citep{cock_sanger_2010}. Despite this convention, the encoding of the Phread score can vary when it is translated to its ASCII representation in the FASTQ file format. For example, the Sanger FASTQ files use ASCII 33–126 to encode Phred qualities from 0 to 93 (i.e. Phred scores with an ASCII offset of 33). A full list of available encoding is available in \ref{}. %ADD FIGURE!

@@

@@

@@

A plethora of open-source tools are available specifically for metagenomic data, both short and long-read data, and several combinations of these tools can be used to characterize the causative agent in a patient's infection in a fraction of the time required by the traditional methods. Alongside, there are several commercial alternatives, such as CLC Genomics Workbench (QIAGEN Bioinformatics), Taxonomer (Flygare et al., 2016) and BaseSpace (Illumina), that offer ready to use complete workflows at the cost of lack of transparency, reproducibility and control in the analysis.
Several steps that can be implemented to ensure the transparency and reproducibility of the chosen workflow. Favouring open-source tools, with clear documentation describing the methodology implemented, and stating the version of the software used and which parameters were used enables the comparison of results. This can be simplified by containerizing all the software tools with one of the many solutions available, like Docker (https://www.docker.com/) or Singularity (Kurtzer et al., 2017). The use of workflow managers, like nextflow (Tommaso et al., 2017) or the Galaxy Project (Afgan et al., 2016), will push reproducibility to the next level by taking advantage of the containerization and scalability, enabling the workflow to be executed with the same parameters in the same conditions in a multitude of different environments. The FlowCraft project (https://github.com/assemblerflow/flowcraft) leverages the combination of Nextflow and docker/singularity containers to assemble, monitor and report scientific pipelines created from the combination of pre-built components, many of them supporting metagenomic analysis.
Additional difficulties of metagenomic data are the overpowering quantities of host DNA that are often sequenced (Couto et al., 2018), making the microbial community close to undetectable, the presence of contaminants, from the bench process to the biota, and the cost associated with this methodology. They account for major caveats and must be made aware of when analysing the data.
The basic strategies for analysing metagenomic data can be simplified in the scheme in Figure 1. One of the biggest challenges when doing metagenomic analysis is differentiating between colonization and infection and to successfully discriminate between a potential pathogen and background microbiota. In the latter, when analysing samples from presumably sterile sites, like CSF and blood, it is safe to assume that all organisms found are of interest. In locations with a microbiota, the use of spiked metagenomic samples as positive control might guide the detection of the possible pathogens by comparing relative abundance between the samples. The inclusion of negative controls is essential for the correct identification of contaminants in the taxonomic results, whether originated from the sample collection, handling or sequencing process. These controls should be processed similarly to the samples and the taxonomic results should be filtered out from the final reports.

\subsection{Quality Assessment and Quality Control}
Quality assessment and control is a basal step to any analysis, and aims to remove and/or filter low quality and low complexity reads, trim adapters, and remove host sequences from the samples’ raw data. There are many tools available but the most commonly used are FastQC (Babraham Bioinformatics) for quality control, followed by Trimmomatic (Bolger et al., 2014) or Cutadapt (Martin, 2011) to trim and/or filter adaptors, low quality and low complexity sequences. For long-read sequencing, tools like NanoPlot and NanoStats (De Coster et al., 2018), Porechop (https://github.com/rrwick/Porechop) and Filtlong (https://github.com/rrwick/Filtlong) can perform basic quality assessment and control, adapter trimming and low quality trimming respectively. 
The removal of host sequences is usually done through mapping. Bowtie2 (Langmead & Salzberg, 2012) is usually the go-tool for this process, specially as pre-build indexes of the human genome are available on-line (https://support.illumina.com/sequencing/sequencing\_software/igenome.html). The same approach can be used to remove contaminants present in the negative control by using it as reference for mapping. Alternatively for long-read, minimap 2 (Li, 2018) can map long-reads to large reference genomes, such as hg19, and the unmapped reads can be subsequently filtered with Samtools (Li et al., 2009).

\subsection{Direct Taxonomic Assignment and Characterization}
An important information that can be retrieved directly from the quality controlled read data is the relative abundance of the detected organisms. Kraken (Wood & Salzberg, 2014) and Bracken (Lu et al., 2017), Midas (Nayfach et al., 2016), Kaiju (Menzel et al., 2016) and MetaPhlAn 2 (Truong et al., 2015) are all examples of taxonomic classifiers for short-read raw sequencing data that provide information on relative abundance. All tools provide a reference database. Kraken offers several pre-built databases constructed from complete bacterial, archaeal, and viral genomes in RefSeq (as of Oct. 18, 2017), known as MiniKraken, that range in size to accommodate limited computational resources, as well as providing the option to build the standard database given that the required computational resources are met. Midas and MetaPhlAn also provide a default pre-built database, with resolution down to strain level. Kaiju differs from the other tools by using a protein reference database but no pre-built version is available, requiring significant resources to build and index the database pre-use. 
The long-read data can be treated as single-end fastq, after proper conversion in the basecalling process. All tools mentioned accommodate classification of single-end files. 
16S classification can also be performed, for validation purposes, by extracting the reads of interest through HMMER (Wheeler & Eddy, 2013) and then perform a traditional OTU analysis with Quiime2 (https://qiime2.org/) or MapSEQ (Rodrigues et al., 2017).  
For virulence gene detection and antimicrobial resistance characterization a mapping approach, with an adequate database, is usually followed, using as reference the Virulence Factors Database (Chen et al., 2016), and ResFinder (Zankari et al., 2012) or CARD (Jia et al., 2017) for antimicrobial resistance gene detection. Besides mapping, other strategies have been applied, like Mash Screen (Ondov et al., 2016), that offer similar results in a faster way.  Similar strategies can be applied to plasmid detection by using the PlasmidFinder (Carattoli et al., 2014) or RefSeq plasmid (O’Leary et al., 2016) databases. The minimap 2 tool (Li, 2018) is a good alternative to map long-read data to any of the resistance and virulence databases mentioned. 
It is possible to genotype the bacterial population in a metagenomic sample, but only for short-read sequencing data.  MetaMLST (Zolfo et al., 2017) reconstructs the MultiLocus Sequence Typing (MLST) loci directly from the sequencing data and provides a pre-built database for the analysis.

@@

\subsection{Virus and Eukaryotes in Metagenomic Analysis}
One of the biggest advantages of using metagenomic methods is the detection of not only bacterial organisms, but also viral and eukaryotic pathogens. 
Besides the limitations inherent to the metagenomic process, the retrieval of viral genomes from clinical samples has added difficulties. The fragments of viral genomes are typically orders of magnitude less abundant, the viral genomes often deviate considerably from reference genomes, and the high intrapopulation viral diversity can lead to ambiguous sequence reconstruction or broken assemblies (Rose et al., 2016). Adding to this, the relatively few viral reference genomes can render classification problematic. For fungi, there’s an underrepresentation of the diversity of this group in databases as it remains understudied compared to bacterial microbiomes (Donovan et al., 2018).
Of the tools mentioned for read classification, Kraken’s MiniKraken and MetaPhlAn 2 databases is the most inclusive, including information of virus, bacteria, human and fungi. None other method of the ones described in this review provide a database as inclusive but many, such as Midas, allow the user to build custom databases although requiring very high computational power. Alternatively, the assembly based method can be implemented followed by an alignment search to a database that includes fungi and viral genomes, such as NCBI’s RefSeq (O’Leary et al., 2016) or GenBank (Benson et al., 2005).

>>>>

%TABLE HERE - Phred and error rate

Since their introduction, Phred scores have become the \textit{de facto} standard for representing sequencing read base qualities \citep{cock_sanger_2010}. Despite this convention, the encoding of the Phread score can vary when it is translated to its ASCII representation in the FASTQ file format. For example, the Sanger FASTQ files use ASCII 33–126 to encode Phred qualities from 0 to 93 (i.e. Phred scores with an ASCII offset of 33). A full list of available encoding is available in \ref{}.

%ADD FIGURE!

@@

\subsection{Direct taxonomic assignment and characterisation}

An important information that can be retrieved directly from the quality controlled read data is the relative abundance of the detected organisms. Kraken (Wood & Salzberg, 2014) and Bracken (Lu et al., 2017), Midas (Nayfach et al., 2016), Kaiju (Menzel et al., 2016) and MetaPhlAn 2 (Truong et al., 2015) are all examples of taxonomic classifiers for short-read raw sequencing data that provide information on relative abundance. All tools provide a reference database. Kraken offers several pre-built databases constructed from complete bacterial, archaeal, and viral genomes in RefSeq (as of Oct. 18, 2017), known as MiniKraken, that range in size to accommodate limited computational resources, as well as providing the option to build the standard database given that the required computational resources are met. Midas and MetaPhlAn also provide a default pre-built database, with resolution down to strain level. Kaiju differs from the other tools by using a protein reference database but no pre-built version is available, requiring significant resources to build and index the database pre-use. 
The long-read data can be treated as single-end fastq, after proper conversion in the basecalling process. All tools mentioned accommodate classification of single-end files. 
16S classification can also be performed, for validation purposes, by extracting the reads of interest through HMMER (Wheeler & Eddy, 2013) and then perform a traditional OTU analysis with Quiime2 (https://qiime2.org/) or MapSEQ (Rodrigues et al., 2017).

For virulence gene detection and antimicrobial resistance characterization a mapping approach, with an adequate database, is usually followed, using as reference the Virulence Factors Database (Chen et al., 2016), and ResFinder (Zankari et al., 2012) or CARD (Jia et al., 2017) for antimicrobial resistance gene detection. Besides mapping, other strategies have been applied, like Mash Screen (Ondov et al., 2016), that offer similar results in a faster way.  Similar strategies can be applied to plasmid detection by using the PlasmidFinder (Carattoli et al., 2014) or RefSeq plasmid (O’Leary et al., 2016) databases. The minimap 2 tool (Li, 2018) is a good alternative to map long-read data to any of the resistance and virulence databases mentioned. 
It is possible to genotype the bacterial population in a metagenomic sample, but only for short-read sequencing data.  MetaMLST (Zolfo et al., 2017) reconstructs the MultiLocus Sequence Typing (MLST) loci directly from the sequencing data and provides a pre-built database for the analysis.

@@

\subsubsection{Genomes through Mapping}

\subsubsection{Genomes through \textit{de novo} assembly}

@@

\subsection{Metataxonomic analysis}

\subsubsection{Amplicon metagenomic analysis}

\subsubsection{Shotgun metagenomic analysis}

% TRASH TEXT
A plethora of open-source tools are available specifically for metagenomic data, both short and long-read data, and several combinations of these tools can be used to characterize the causative agent in a patient's infection in a fraction of the time required by the traditional methods. Alongside, there are several commercial alternatives, such as CLC Genomics Workbench (QIAGEN Bioinformatics), Taxonomer (Flygare et al., 2016) and BaseSpace (Illumina), that offer ready to use complete workflows at the cost of lack of transparency, reproducibility and control in the analysis.
Several steps that can be implemented to ensure the transparency and reproducibility of the chosen workflow. Favouring open-source tools, with clear documentation describing the methodology implemented, and stating the version of the software used and which parameters were used enables the comparison of results. This can be simplified by containerizing all the software tools with one of the many solutions available, like Docker (https://www.docker.com/) or Singularity (Kurtzer et al., 2017). The use of workflow managers, like nextflow (Tommaso et al., 2017) or the Galaxy Project (Afgan et al., 2016), will push reproducibility to the next level by taking advantage of the containerization and scalability, enabling the workflow to be executed with the same parameters in the same conditions in a multitude of different environments. The FlowCraft project (https://github.com/assemblerflow/flowcraft) leverages the combination of Nextflow and docker/singularity containers to assemble, monitor and report scientific pipelines created from the combination of pre-built components, many of them supporting metagenomic analysis.
Additional difficulties of metagenomic data are the overpowering quantities of host DNA that are often sequenced (Couto et al., 2018), making the microbial community close to undetectable, the presence of contaminants, from the bench process to the biota, and the cost associated with this methodology. They account for major caveats and must be made aware of when analysing the data.
The basic strategies for analysing metagenomic data can be simplified in the scheme in Figure 1. One of the biggest challenges when doing metagenomic analysis is differentiating between colonization and infection and to successfully discriminate between a potential pathogen and background microbiota. In the latter, when analysing samples from presumably sterile sites, like CSF and blood, it is safe to assume that all organisms found are of interest. In locations with a microbiota, the use of spiked metagenomic samples as positive control might guide the detection of the possible pathogens by comparing relative abundance between the samples. The inclusion of negative controls is essential for the correct identification of contaminants in the taxonomic results, whether originated from the sample collection, handling or sequencing process. These controls should be processed similarly to the samples and the taxonomic results should be filtered out from the final reports.

@@

\subsection{Virus in Metagenomic Analysis}
One of the biggest advantages of using metagenomic methods is the detection of not only bacterial organisms, but also viral and eukaryotic pathogens. 
Besides the limitations inherent to the metagenomic process, the retrieval of viral genomes from clinical samples has added difficulties. The fragments of viral genomes are typically orders of magnitude less abundant, the viral genomes often deviate considerably from reference genomes, and the high intrapopulation viral diversity can lead to ambiguous sequence reconstruction or broken assemblies (Rose et al., 2016). Adding to this, the relatively few viral reference genomes can render classification problematic. For fungi, there’s an underrepresentation of the diversity of this group in databases as it remains understudied compared to bacterial microbiomes (Donovan et al., 2018).
Of the tools mentioned for read classification, Kraken’s MiniKraken and MetaPhlAn 2 databases is the most inclusive, including information of virus, bacteria, human and fungi. None other method of the ones described in this review provide a database as inclusive but many, such as Midas, allow the user to build custom databases although requiring very high computational power. Alternatively, the assembly based method can be implemented followed by an alignment search to a database that includes fungi and viral genomes, such as NCBI’s RefSeq (O’Leary et al., 2016) or GenBank (Benson et al., 2005).
